Unbalanced Mallows Models for Optimizing Expensive Black-Box Permutation Problems

### 2021-02-04 paper/main.tex
tuples

\begin{algorithm}[t]
 \caption{UMM: Unbalanced Mallows Model Algorithm}
 \label{alg:umm}
  \begin{algorithmic}[1]
    \Require $\minit$: number of initial permutations, $\FEmax$: total budget,\newline\hspace*{5ex} \manuel{$\rho$: learning rate,}{$r_1, r_2$: parameters controlling the learning rate} 
    % \\ ?, $\FEmax$: budget of evaluations,\newline $S$: initial permutations, $F$: their fitness values
    \State $S \assign \{\}$
    \For{$i \assign 0$ \textbf{to} $\minit$}
    \State $\sigma \assign \text{generate uniformly at random}$
    \State $S \assign S \cup \tuple{\sigma, f(\sigma)}$\Comment{Evaluate it}
    \EndFor
    %\State $S \assign $ generate  $\minit$ permutations uniformly at random
    %\State $F \assign \{ f(\sigma) \mid \sigma \in S\}$ \Comment{Evaluate them}
    \For{$i \assign \minit$ \textbf{to} $\FEmax$ \text{evaluations}}
    \State $\hat\sigma_0 \assign \text{uBorda}(S, F,\rho)$\label{line:uborda}\Comment{Learning Step}
    \State Initialize (first iteration) or decrease  $\theta$\label{line:theta}
    \Statex \Comment{See \emph{Sampling Step} and Eq.~\ref{eq:expectation}}
   % \State $\Sigma \assign$ sample $\beta$ permutations from  $MM(\hat\sigma_0, \theta)$\label{line:sample}
     \State $\sigma \assign$ sample a permutation from  $MM(\hat\sigma_0, \theta)$\label{line:sample}
    % \Comment{Generate $\beta$ permutations}
    %\State $\sigma\assign  \argmax_{\pi'\in \Sigma}\min_{ \pi \in S} d(\pi', \pi)   $ \label{line:distance}
    %\Statex \Comment{Select the farthest from $S$}
   % \State $\sigma \assign$ sampled from  $MM(\hat\sigma_0, \theta)$\label{line:sample}
    \State $S \assign S \cup \tuple{\sigma, f(\sigma)}$\Comment{Evaluate it}
    \EndFor 
    \State \Return $\tuple{\sigma^\ast, f(\sigma^\ast)}\assign \argmin_{\tuple{\sigma, f(\sigma)}\in S} f(\sigma)$
  %   \Require $n$: ?, $m$: budget of evaluations,\newline $S$: initial permutations, $F$: their fitness values
  %   \For{$i \assign 1$ \textbf{to} $m$ \text{evaluations}}
  %   \State $\hat\sigma_0 \assign \text{uBorda}(S, F,\rho)$
  %   \State linear decrease on $\theta$
  %   \State $\sigma \assign$ sample from a $MM(\hat\sigma_0, \theta)$
  %   \State $S \assign S \cup \{\sigma\}$
  %   \State $F \assign F \cup \{f(\sigma)\}$
  %   \EndFor 
  %   \State \Return $S$, $F$
 \end{algorithmic}
\end{algorithm}

@@

\subsection{Learning step}
%
At each iteration, we start by fitting a MM to the sample $S$, which implies
estimating $\hat\sigma_0$ and $\theta$. Parameter $\theta$ is updated
deterministically at the sampling step.  The parameter $\sigma_0$, however, is
learned from the current sample $S$ by using the uBorda algorithm with a weight
$w(\sigma)=\rho^{f(\sigma)}$ (Section~\ref{sec:uborda}). The learning rate is
controlled by parameter $0 < \rho < 1$. \MANUEL{Don't we have to say something about $f(\sigma) > 1$?}
In UMM, the value of $\rho$ depends on the set of true fitness evaluations $F$
and it is set at each iteration such that the largest $100r_1$\% of the mass of
the weights is concentrated in the best $100r_2$\% of the solutions in $S$:
%
\MANUEL{Reviewer 3:  Justify the form of the formula used to learn the "learning rate" $\rho$ (equation 4). Why this particular approach was chosen?}
%
\begin{equation}
  \begin{split}
  r_1 \sum_{\sigma\in S}\rho^{f(\sigma)} =  \sum_{\sigma'\in S'}\rho^{f(\sigma')} \quad\text{where}\; &|S'| = r_2|S| \\[-1.2em]
  &\land \forall \sigma'\in S', \sigma\in S\setminus S' : f(\sigma')\leq f(\sigma).
  \end{split}
\end{equation}
%
i.e., $S'$ contains the $r_2|S|$ elements of $S$ with the best corresponding fitness values in $F$. %The values of $r_1, r_2 \in (0, 1)$ are user-defined and control the balance between exploration and intensification. %: extreme different values bias the sampling towards the best solutions found so far, which may lead to quick stagnation, whereas values (0.5,0.5) make the sampling more uniform, thus slowing down the learning.

>>>>

\begin{algorithm}[t]
 \caption{UMM: Unbalanced Mallows Model Algorithm}
 \label{alg:umm}
  \begin{algorithmic}[1]
    \Require $\minit$: number of initial permutations, $\FEmax$: total budget,\newline\hspace*{5ex} $r_1, r_2$: parameters controlling the learning rate 
    % \\ ?, $\FEmax$: budget of evaluations,\newline $S$: initial permutations, $F$: their fitness values
    \State $S \assign \{\}$
    \For{$i \assign 0$ \textbf{to} $\minit$}
    \State $\sigma \assign \text{generate uniformly at random}$
    \State $S \assign S \cup \tuple{\sigma, f(\sigma)}$\Comment{Evaluate it}
    \EndFor
    %\State $S \assign $ generate  $\minit$ permutations uniformly at random
    %\State $F \assign \{ f(\sigma) \mid \sigma \in S\}$ \Comment{Evaluate them}
    \For{$i \assign \minit$ \textbf{to} $\FEmax$ \text{evaluations}}
    \State $\hat\sigma_0 \assign \text{uBorda}(S, F,\rho)$\label{line:uborda}\Comment{Learning Step}
    \State Initialize (first iteration) or decrease  $\theta$\label{line:theta}
    \Statex \Comment{See \emph{Sampling Step} and Eq.~\ref{eq:expectation}}
   % \State $\Sigma \assign$ sample $\beta$ permutations from  $MM(\hat\sigma_0, \theta)$\label{line:sample}
     \State $\sigma \assign$ sample a permutation from  $MM(\hat\sigma_0, \theta)$\label{line:sample}
    % \Comment{Generate $\beta$ permutations}
    %\State $\sigma\assign  \argmax_{\pi'\in \Sigma}\min_{ \pi \in S} d(\pi', \pi)   $ \label{line:distance}
    %\Statex \Comment{Select the farthest from $S$}
   % \State $\sigma \assign$ sampled from  $MM(\hat\sigma_0, \theta)$\label{line:sample}
    \State $S \assign S \cup \tuple{\sigma, f(\sigma)}$\Comment{Evaluate it}
    \EndFor 
    \State \Return $\tuple{\sigma^\ast, f(\sigma^\ast)}\assign \argmin_{\tuple{\sigma, f(\sigma)}\in S} f(\sigma)$
  %   \Require $n$: ?, $m$: budget of evaluations,\newline $S$: initial permutations, $F$: their fitness values
  %   \For{$i \assign 1$ \textbf{to} $m$ \text{evaluations}}
  %   \State $\hat\sigma_0 \assign \text{uBorda}(S, F,\rho)$
  %   \State linear decrease on $\theta$
  %   \State $\sigma \assign$ sample from a $MM(\hat\sigma_0, \theta)$
  %   \State $S \assign S \cup \{\sigma\}$
  %   \State $F \assign F \cup \{f(\sigma)\}$
  %   \EndFor 
  %   \State \Return $S$, $F$
 \end{algorithmic}
\end{algorithm}

@@

\subsection{Learning step}
%
At each iteration, we start by fitting a MM to the sample $S$, which implies
estimating $\hat\sigma_0$ and $\theta$. Parameter $\theta$ is updated
deterministically at the sampling step.  The parameter $\sigma_0$, however, is
learned from the current sample $S$ by using the uBorda algorithm with a weight
$w(\sigma)=\rho^{f(\sigma)}$ (Section~\ref{sec:uborda}). The learning rate is
controlled by parameter $0 < \rho < 1$. \MANUEL{Don't we have to say something about $f(\sigma) > 1$?}
In UMM, the value of $\rho$ depends on the set of true fitness evaluations $F$
and it is set at each iteration such that the largest $100r_1$\% of the mass of
the weights is concentrated in the best $100r_2$\% of the solutions in $S$:
%
\MANUEL{Reviewer 3:  Justify the form of the formula used to learn the "learning rate" $\rho$ (equation 4). Why this particular approach was chosen?}
%
\begin{equation}
%  \begin{split}
    r_1 \cdot \sum_{\tuple{\sigma,f(\sigma)}\in S}\rho^{f(\sigma)} =  \sum_{\tuple{\sigma', f(\sigma')}\in S'}\rho^{f(\sigma')}  %\\
  %    & \text{where}\;|S'| = r_2|S| \land \forall \sigma'\in S', \sigma\in S\setminus S' : f(\sigma')\leq f(\sigma).
%  \end{split}
\end{equation}
%
where $|S'| = r_2|S| \land \forall \tuple{\sigma',f(\sigma')}\in S', \tuple{\sigma,f(\sigma)}\in S\setminus S' : f(\sigma')\leq f(\sigma)$, that is, $S'$ contains the $r_2|S|$ elements of $S$ with the best  fitness values. %The values of $r_1, r_2 \in (0, 1)$ are user-defined and control the balance between exploration and intensification. %: extreme different values bias the sampling towards the best solutions found so far, which may lead to quick stagnation, whereas values (0.5,0.5) make the sampling more uniform, thus slowing down the learning.
